### 2021-07-26 ch/1-introduction.tex
Finish

Fairness metrics require a fairness \emph{definition},
and the fairness definition with the clearest interpretation in the described setup of label bias and sampling bias is \acfi{DP},
also called \emph{statistical parity} or \emph{independence}.
It demands that the predictions $\hat{y}$ be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them, the difference and the ratio of the terms on the two sides of the equation.

@@

\section{Relations to other fields and clarification of terms}
% \subsection{aka: Isn't this thesis actually about XY instead of fairness?}
The thesis is principally written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

>>>>

Fairness metrics require a fairness \emph{definition},
and the fairness definition with the clearest interpretation in the described setup of label bias and sampling bias is \acfi{DP},
also called \emph{statistical parity} or \emph{independence}.
It demands that the predictions $\hat{y}$\, be independent of the sensitive attribute \(s\).
So, for binary $s$ and $y$\,:
\begin{align}
  P(\hat{y}=1|s=0) &= P(\hat{y}=1|s=1)~.
  \label{eq:dp-def}
\end{align}
There are multiple \ac{DP} \emph{metrics} which track how close the predictions are to satisfying the equality,
among them, the difference and the ratio of the terms on the two sides of the equation.

@@

\section{Relations to other fields and clarification of terms}
% \subsection{aka: Isn't this thesis actually about XY instead of fairness?}
This thesis is principally written from the perspective of algorithmic fairness,
but it touches on other fields as well, like \emph{domain shift} and \emph{causality}.
Furthermore, the thesis makes use of concepts like \emph{transferable representations}
and \emph{interpretability}.
I summarise the areas here and discuss their relevance to the presented work.

### 2021-07-26 ch/2-related-work.tex
Clean up
Try to improve taxonomy figure
Finish

      [data de-biasing

@@

        [fair repre-\\sentations

@@

          [adversarial\\networks

@@

          [modified\\sampling]

@@

              [change of\\optimisation target, draw=blue]

@@

          [loss\\reweighting]

@@

@@

    The three categories with blue borders are those categories
    to which the proposed models in this thesis belong (see \chapref{ch:content} for details).

@@

\Figref{fig:fairness-taxonomy} provides a high-level overview
of one possible categorisation of the methods presented here.
It is meant as a rough guide through the chapter.

>>>>

      [data de-biasing, line width = 2pt

@@

        [additional\\synthetic data]
          [fair representations

@@

            [adversarial networks

@@

        [modified sampling]

@@

              [change of optim-\\isation target, draw=blue]

@@

@@

        [loss reweighting]
      ]]

@@

    The three categories with blue borders correspond to those categories
    under which the proposed models in this thesis fall (see \chapref{ch:content} for details).

@@

\Figref{fig:fairness-taxonomy} provides a high-level overview
of one possible categorisation of the methods presented here.
It can serve as a map for navigating through this chapter.
% It is meant as a rough guide through the chapter.

### 2021-07-26 ch/7-discussion.tex
Finish

The remainder of this chapter is split into three parts:
In \secref{sec:ch7-limitations},
I discuss some of the limitations of the presented methods to keep in mind,
and what the goal of the applications of the methods should be.
\Secref{sec:potential-extentions} is about ways in which the work could be extended,
and \secref{sec:broader-perspective} discusses the broader perspective of this topic,
including a wider discussion of the future of the field.

@@

% ============ price of fairness ==========
It also should be mentioned here that compared to training straightforward, bias-unaware classifiers,
the sampling-bias-targeting methods (\ie, the second and the third method)
incur additional costs in terms of computing resources, training time and human labour.
This label-bias-targeting method (\ie, the first one) only impacts the computation of the loss
and does not require additional training data,
so it has a negligible impact training cost.
The second method is the most costly
as the training of \acp{INN} requires a large amount of memory and compute
because \acp{INN} only \emph{transform} the (very high-dimensional) inputs
and never perform anything like the lossy compression that other neural network do.
Furthermore, the method requires additional data with \(s\) labels.
Still, if the additional training cost enables the use of data
that was previously unusable due to strong spurious correlations,
then it can still be worthwhile,
and it is also important to keep in mind that a model can be trained once
and then used millions of times.
This has to be decided on a case-by-case basis,
and unfortunately means that these de-biasing methods will likely not be use prophylactically,
but only if there is an explicit expectation of harm otherwise.
% ==============

@@

% ====== why even work with bad data? =====
Now, one possible objection here is:
if those datasets are of such poor quality,
then maybe we just should not train any \ac{ML} model on these and should not use them to make automated decisions.
While this question is mostly beyond the scope of this document, let me offer some thoughts on this:
It is true that even after the application of de-biasing techniques,
the resulting models still should not be fully trusted,
but they can still ease the burden of tedious manual labour;
similar to an email spam filter which is not perfect, but still very useful.
Or, put another way, it is always important to check what the realistic alternative is;
we should not compare a model to a non-existing perfect ideal, but to the actual solution that would be used instead.
One could imagine a hybrid approach where an automated system makes preliminary decisions,
but random samples are reviewed by humans and decisions can always be challenged.
Ideally, the model itself would tell us about decisions it is uncertain about.

@@

% ======== data augmentation ========
A topic that was left out of the thesis is \emph{data augmentation},
which refers to the practice of modifying copies of the samples in the training set
with pre-defined transformations, which do not change the ``semantic content'',
such that more samples are available for training.
For image data, typical transformations are rotations, crops, mirroring, Gaussian noise, and slight colour modifications.
If chosen right, data augmentation alone can solve the problems presented in this thesis:
grey-scaling is a transformation that is sometimes used for image data,
and works quite well on the Coloured MNIST problem
(though not perfectly well, because even as shades of grey, the colours are distinguishable).
However, this only works because the spurious feature is so simply here
(after all, Coloured MNIST is mostly intended to be a simple toy dataset).
Grey-scaling would not help with real-world datasets like CelebA.
Furthermore, even if a simple transformation would be able to remove the spurious feature,
knowing \emph{which} transformation will accomplish this is not always as obvious as with Coloured MNIST,
and may require special domain knowledge (like signal processing for audio data).
Thus, the goal for the presented methods here was to make them applicable do any kind of data,
without requiring knowledge of the specific details of how the bias manifests itself in the data,
and without requiring knowledge of which data augmentations can be applied.
However, this is not a recommendation \emph{against} data augmentations;
if you have reason to believe that augmentations will help,
there is no reason not to use them.
% ==============

\section{Potential extensions}\label{sec:potential-extentions}
%
% ======= potential technical improvements ======
% There are also certain technical improvements that could be made.
In general, it would be desirable to have more theoretical bounds on performance.
This likely involves specifying the requirements for the algorithms more precisely.
% which is also useful information for potential users.
The method based on target labels would furthermore benefit from better-calibrated probability outputs.
\acf{GP} classifiers show generally good calibration,
but cannot be easily applied in domains where deep neural networks are used.
On the other hand, neural networks are known to be overconfident
\citep[especially when using ReLU activations;][]{hein2019relu},
so to apply the proposed method to images directly is not straightforward.
The other two proposed methods would benefit from improved adversarial training procedures.
Training these models is not straightforward as the losses have to be balanced
and in some cases, the update schedule needs to be changed.
Both the calibration of neural networks and the stability of adversarial training
are issues that are widely recognised in the \ac{ML} community,
so there is hope that progress will be made on these.
% ==============

@@

% ======== what can we do if we don't know there is a bias? ========
A potential -- very ambitious -- extension would be
to try to learn to automatically to detect biases:
\ie an algorithm that can recognise biases without prompting a human operator.
The question is then, where can an \ac{ML} algorithm learn about all possible biases?
This could potentially be done with models like GPT-3 \citep{brown2020language},
which are trained on enormous amounts of text data,
such that they conceivably have learned a lot about what human writers consider to be biases.
There is then perhaps a way to extract what such a model knows about biases
(it could be as easy as a text prompt which asks ``is it okay to discriminate based on gender''),
which would allow us to check whether those biases exist in our data.
Of course, this assumes that humans agree what biases are --
otherwise the model might be very unreliable --
which seems like quite a risky assumption to make.
There are also other potential problems with unsupervised models like GPT-3;
some of which are discussed below.
% ==============

\section{Broader perspective}\label{sec:broader-perspective}
%
% ====== human-machine interface aspect =======
An important issue is the communication between human and machine.
The methods presented in this thesis have all strived to make it easy for human operators to understand what is going on:
the method in \chapref{ch:paper1} is configured with simple statistics;
the method in \chapref{ch:paper2} produces invariant images that can be inspected;
and the method in \chapref{ch:paper3} has the same capability
(though it is not part of the core functionality).
However, this is only a beginning.
It is still not easy to notice that a given dataset has problems,
and machines on their own cannot notice the problem,
because they do not (yet) have a complete understanding about what it is that humans care about;
human preferences are incredibly complex \citep{yudkowsky2011complex} and hard to predict for today's algorithms.
Thus, it is important that machines get feedback early and often,
to keep them aligned with human goals.
% (The aforementioned \citet{stiennon2020learning} is arguably an example of this.)
Applied to the problem of dataset bias,
this could mean visualising correlations in the dataset,
or routinely producing invariant representations to show what the network thinks it is meant to learn.
% ==============

% ========= unsupervised learning =======
One area of machine learning that has recently seen increased interest is unsupervised learning,
and the latter two chapters make use of it to some degree.
The exciting promise of unsupervised learning in general is
that labour-intensive labelling is not needed and so vast amounts of existing, unlabelled data can be put to good use.
One could ask the question whether bias-correcting methods are still needed, with access to so much data.
It could be that, while the data is certainly not perfect,
there is so much of it that the biased parts ``cancel each other out''.
However, recent investigations of the GPT models \citep{radford2018improving,radford2019language,brown2020language} do not seem to support this \citep{khalifa2021distributional}.
One reason for this might be the way these models are trained at the moment:
they maximise the probability assigned to the next token.
Thus, such a model has to account for the wide array of human opinions and assign a non-zero probability to all of them.
So, when asked to summarise a text \citep{stiennon2020learning}, GPT does not give the \emph{best} summary;
rather, it gives a summary that an average person might have written.
However, with the help of a very high-quality labelled dataset (that was expensive to create),
GPT could be finetuned to actually produce very good summaries.
I suspect this pattern of learning the basics in an unsupervised fashion,
and then finetuning with high-quality labels, will continue in the future.
With these massive models,
it is, more than ever, crucial to build tools to make the biases within the models transparent.
Given the black-box nature of neural networks, this represents a major challenge.
% When examining these models trained on massive datasets through the lens of the dataset biases discussed in \chapref{ch:related-work},
% we encounter a novel problem:
% somewhere in the model,
% (an approximation of) the fundamental structures of reality are likely represented,
% but we cannot easily access this.
% ==============

% ======== is it fair that AI has to be fair, if humans tend to be unfair anyhow? ========
%   - yes, because there is no excuse to make something artificially biased, and also, an AI might affect many more people than a single person
%   - it's like asking: isn't it unfair that our president should have a better character than the average person?
It was mentioned before that an \ac{ML} model should be judged by how useful it is,
how it stacks up to realistic alternatives, and not if it is perfect.
However, one has to be careful not to take this philosophy too far.
Namely, one should avoid giving up and saying,
``why should AI need to be fair if humans tend to be bias anyhow?''
The faults of humans has little to do with the question of what we expect of machines.
% and we should certainly not go out of our way to encode the same biases in our creations.
If one of our machines will affect many of our fellow humans,
then I think we would not want it to be harmful.
Should we make the machine artificially biased
so that humans do not need to feel too guilty about their own biases?
I believe this to be folly.
We should always strive to do the best we can
and that means making machines as unbiased as possible.
% ==============

>>>>

The remainder of this chapter is split into three parts:
In \secref{sec:ch7-limitations},
I discuss some of the limitations of the presented methods to keep in mind,
and the goals that can be achieved by applying the methods.
\Secref{sec:potential-extentions} is about ways in which the work could be extended,
and \secref{sec:broader-perspective} discusses the broader perspective of this topic,
including a wider discussion of the future of the field.

@@

% ============ price of fairness ==========
It also should be mentioned here that compared to training straightforward, bias-unaware classifiers,
the sampling-bias-targeting methods (\ie, the second and the third method)
incur additional costs in terms of computing resources, training time and human labour.
The label-bias-targeting method (\ie, the first one) only impacts the computation of the loss
and does not require additional training data,
so it has a negligible impact on training cost.
The second method is the most costly
as the training of \acp{INN} requires a large amount of memory and compute
because \acp{INN} only \emph{transform} the (very high-dimensional) inputs
and never perform anything like the lossy compression that other neural network do.
Furthermore, the method requires additional data with \(s\) labels.
Still, if the additional training cost enables the use of data
that was previously unusable due to strong spurious correlations,
then it can still be worthwhile,
and it is also important to keep in mind that a model can be trained once
and then used millions of times.
This has to be decided on a case-by-case basis,
and unfortunately means that these de-biasing methods will likely not be use prophylactically,
but only if there is an explicit expectation of harm otherwise.
% ==============

@@

% ====== why even work with bad data? =====
Now, one possible objection here is:
if those datasets are of such poor quality,
then maybe we should not train any \ac{ML} model on these and should not use them to make automated decisions.
While this question is mostly beyond the scope of this document, let me offer some thoughts on this:
It is true that even after the application of de-biasing techniques,
the resulting models still should not be fully trusted,
but they can still ease the burden of tedious manual labour;
similar to an email spam filter which is not perfect, but still very useful.
Or, put another way, it is always important to check what the realistic alternative is;
we should not compare a model to a non-existing perfect ideal, but to the actual solution that would be used instead.
One could imagine a hybrid approach where an automated system makes preliminary decisions,
but random samples are reviewed by humans and decisions can always be challenged.
Ideally, the model itself would tell us about decisions it is uncertain about.

@@

% ======== data augmentation ========
A topic that was left out of the thesis is \emph{data augmentation},
which refers to the practice of modifying copies of the samples in the training set
with pre-defined transformations, which do not change the ``semantic content'',
such that more samples are available for training.
For image data, typical transformations are rotations, crops, mirroring, Gaussian noise, and slight colour modifications.
If chosen right, data augmentation alone can solve the problems presented in this thesis:
grey-scaling is a transformation that is sometimes used for image data,
and works quite well on the Coloured MNIST problem
(though not perfectly well, because even as shades of grey, the colours are distinguishable).
However, this only works because the spurious feature is so simply here
(after all, Coloured MNIST is mostly intended to be a simple toy dataset).
Grey-scaling would not help with real-world datasets like CelebA.
Furthermore, even if a simple transformation would be able to remove the spurious feature,
knowing \emph{which} transformation will accomplish this is not always as obvious as with Coloured MNIST,
and may require special domain knowledge (like signal processing for audio data).
Thus, the goal for the presented methods here was to make them applicable to any kind of data,
without requiring knowledge of the specific details of how the bias manifests itself in the data,
and without requiring knowledge of which data augmentations can be applied.
However, this is not a recommendation \emph{against} data augmentations;
if you have reason to believe that augmentations will help,
there is no reason not to use them.
% ==============

\section{Potential extensions}\label{sec:potential-extentions}
%
% ======= potential technical improvements ======
% There are also certain technical improvements that could be made.
In general, it would be desirable to have more theoretical bounds on performance.
This would likely involve specifying the requirements for the algorithms more precisely.
% which is also useful information for potential users.
The method based on target labels would furthermore benefit from better-calibrated probability outputs.
\acf{GP} classifiers show generally good calibration,
but cannot be easily applied in domains where deep neural networks are used.
On the other hand, neural networks are known to be overconfident
\citep[especially when using ReLU activations;][]{hein2019relu},
so to apply the proposed method to images directly is not straightforward.
The other two proposed methods would benefit from improved adversarial training procedures.
Training these models is not straightforward as the losses have to be balanced
and in some cases, the update schedule needs to be changed.
Both the calibration of neural networks and the stability of adversarial training
are issues that are widely recognised in the \ac{ML} community,
so there is hope that progress will be made on these.
% ==============

@@

% ======== what can we do if we don't know there is a bias? ========
A potential -- very ambitious -- extension would be
to try to learn to automatically detect biases:
\ie\ an algorithm that can recognise biases without asking a human operator.
The question is then, where can an \ac{ML} algorithm learn about all possible biases?
Given the complexity of human values \citep{yudkowsky2011complex},
it will not be possible to compress knowledge of all biases down to a compact representation.
Essentially, the only way is to learn them all individually, one by one.
This could potentially be done with models like GPT-3 \citep{brown2020language},
which are trained on enormous amounts of text data,
such that they conceivably have learned a lot about what human writers consider to be biases.
There is then perhaps a way to extract what such a model knows about biases
(it could be as easy as a text prompt which asks ``is it okay to discriminate based on gender?''),
which would allow us to check whether those biases exist in our data.
Of course, this assumes that humans agree amongst themselves what biases are --
otherwise the model might be very unreliable --
which seems like quite a risky assumption to make.
Furthermore, prevalent ethics have change over the course of human history,
and are likely to change in the future as well.
The ideal solution would be to predict which ethics humans will adopt in the future,
after they thought about it even more,
which is the goal of \emph{coherent extrapolated volition} \citep{tarleton2010coherent},
but has large technical hurdles.
There are also other potential problems with unsupervised models like GPT-3;
some of which are discussed below.
% ==============

\section{Broader perspective}\label{sec:broader-perspective}
%
% ====== human-machine interface aspect =======
An important issue is the communication between human and machine.
The methods presented in this thesis have all strived to make it easy for human operators to understand what is going on:
the method in \chapref{ch:paper1} is configured with simple statistics;
the method in \chapref{ch:paper2} produces invariant images that can be inspected;
and the method in \chapref{ch:paper3} has the same capability
(though it is not part of the core functionality).
However, this is only a beginning.
It is still not easy to notice that a given dataset has problems,
and \emph{currently}, machines on their own cannot notice the problem (see above).
Thus, it is important that machines get feedback early and often,
to keep them aligned with human goals.
% (The aforementioned \citet{stiennon2020learning} is arguably an example of this.)
Applied to the problem of dataset bias,
this could mean visualising correlations in the dataset,
or routinely producing invariant representations to show what the network thinks it is meant to learn.
% ==============

% ========= unsupervised learning =======
One area of machine learning that has recently seen increased interest is unsupervised learning,
and the latter two chapters make use of it to some degree.
The exciting promise of unsupervised learning in general is
that labour-intensive labelling is not needed and so vast amounts of existing, unlabelled data can be put to good use.
One could ask the question whether bias-correcting methods are still needed, with access to so much data.
It could be that, while the data is certainly not perfect,
there is so much of it that the biased parts ``cancel each other out''.
However, recent investigations into the GPT models \citep{radford2018improving,radford2019language,brown2020language} do not seem to support this \citep{khalifa2021distributional}.
One reason for this might be the way these models are trained at the moment:
they maximise the probability assigned to the next token.
Thus, such a model has to account for the wide array of human opinions and assign a non-zero probability to all of them.
So, when asked to summarise a text \citep{stiennon2020learning}, GPT does not give the \emph{best} summary;
rather, it gives a summary that an average person might have written.
However, with the help of a very high-quality labelled dataset (that was expensive to create),
it was possible to fintetune GPT to actually produce very good summaries.
I suspect this pattern of learning the basics in an unsupervised fashion,
and then finetuning with high-quality labels, will continue in the future.
With these massive models,
it is, more than ever, crucial to build tools to make the biases within the models transparent.
Given the black-box nature of neural networks, this represents a major challenge.
% When examining these models trained on massive datasets through the lens of the dataset biases discussed in \chapref{ch:related-work},
% we encounter a novel problem:
% somewhere in the model,
% (an approximation of) the fundamental structures of reality are likely represented,
% but we cannot easily access this.
% ==============

% ======== is it fair that AI has to be fair, if humans tend to be unfair anyhow? ========
%   - yes, because there is no excuse to make something artificially biased, and also, an AI might affect many more people than a single person
%   - it's like asking: isn't it unfair that our president should have a better character than the average person?
It was mentioned before that an \ac{ML} model should be judged by how useful it is,
how it stacks up to realistic alternatives, and not whether it is perfect.
However, one has to be careful not to take this philosophy too far.
Namely, one should avoid giving up and saying,
``why should AI need to be fair if humans tend to be biased anyhow?''
The faults of humans has little to do with the question of what we expect of machines.
% and we should certainly not go out of our way to encode the same biases in our creations.
If one of our machines will affect many of our fellow humans,
then I think we would not want it to be harmful.
Should we make the machine artificially biased
so that humans do not need to feel too guilty about their own biases?
I believe this to be folly.
We should always strive to do the best we can
and that means making machines as unbiased as possible.
% ==============

### 2021-07-26 paper3/appendix.tex
Finish

For the Coloured MNIST and CelebA datasets, the baseline \texttt{CNN}, \texttt{DRO}, and \texttt{LfF} (in the case of the former) models use an architecture identical to that of the encoder with two exceptions: 1) max-pooling being used for spatial downsampling instead of strided convolutions; 2) the final convolutional layer is followed by a global average pooling layer followed by a fully-connected classification layer. For evaluating our method, we simply train a linear classifier on top of $z_y$; this is sufficient due to linear-separability being enforced during training by the $y$-predictor.
For the Adult Income dataset, we use an MLP made up of a single hidden layer 35 units in size, followed by a SELU activation \citep{klambauer2017self}, as both both the downstream classifier for our method, and as the network architecture of the baselines. 
All baselines and downstream classifiers alike were trained for $60$ epochs with a learning rate of $1 \times 10^{-3}$ and a batch size of $256$.

Since, by design, we do not have labels for all subgroups the model will be tested on, and bias against these missing subgroups is what we aim to avoid, properly validating, and thus conducting hyperparameter selection for models generally, is not straightforward.
We can use estimates of the mutual information between the learned-representation and $s$ and $y$ (which we wish to minimise w.r.t.\ to the former, maximise w.r.t.\ the latter) to guide the process, though optimising the model w.r.t.\ to these metrics obtained from only the training set does not guarantee generalisation to the missing subgroups.
We can, however, additionally measure the entropy of the predictions on the encoded test set and seek to maximise it across all samples, or alternatively train a discriminator of the same kind used for distribution matching as a measure the shift in the latent space between datasets.
We use the latter approach (considering, the learned distance between subspace distributions, accuracy, and reconstruction loss) to inform an extensive grid-search over the hyperparameter space of our model.

>>>>

For the Coloured MNIST and CelebA datasets, the baseline \texttt{CNN}, \texttt{DRO}, and \texttt{LfF} (in the case of the former) models use an architecture identical to that of the encoder with two exceptions: 1) max-pooling being used for spatial downsampling instead of strided convolutions; 2) the final convolutional layer is followed by a global average pooling layer followed by a fully-connected classification layer. For evaluating our method, we simply train a linear classifier on top of $z_y$; this is sufficient due to linear-separability being enforced during training by the $y$-predictor.
For the Adult Income dataset, we use an \ac{MLP} made up of a single hidden layer -- 35 units in size -- followed by a SELU activation \citep{klambauer2017self}, as both both the downstream classifier for our method, and as the network architecture of the baselines. 
All baselines and downstream classifiers alike were trained for $60$ epochs with a learning rate of $1 \times 10^{-3}$ and a batch size of $256$.

Since, by design, we do not have labels for all subgroups the model will be tested on, and bias against these missing subgroups is what we aim to avoid, properly validating, and thus conducting hyperparameter selection for models generally, is not straightforward.
We can use estimates of the mutual information between the learned-representation and $s$ and $y$ (which we wish to minimise w.r.t.\ to the former, maximise w.r.t.\ the latter) to guide the process, though optimising the model w.r.t.\ these metrics obtained from only the training set does not guarantee generalisation to the missing subgroups.
We can, however, additionally measure the entropy of the predictions on the encoded test set and seek to maximise it across all samples, or alternatively train a discriminator of the same kind used for distribution matching as a measure the shift in the latent space between datasets.
We use the latter approach (considering, the learned distance between subspace distributions, accuracy, and reconstruction loss) to inform an extensive grid-search over the hyperparameter space of our model.
